### 2024-08-13 skripsi/bab1.tex
chapter 1 outline, WIP chapter 1: machine learning in physics history

\chapter{INTRODUCTION}
\section{Background}
\label{sec:background} %pelabelan ini opsional, biar bisa di klik kalo mengarahkan ke section tertentu.

>>>>

\chapter{INTRODUCTION}
\section{Background}
\label{sec:background} %pelabelan ini opsional, biar bisa di klik kalo mengarahkan ke section tertentu.
\noindent Partial differential equations (PDE) are a common tool widely used in the modern scientific understanding and many engineering processes. This is because many systems can be described by the way they change and often this can be more intuitive. As an example, when cooking with a frying pan on a stove top that is too small, notice that parts of the pan closer to the burner or heating element would heat up faster and stay hotter compared to the edge even after it stops heating up any further and reach a stable temperature. However, a more useful way to describe the system is using the heat equation. For a temperature function $u\left(\vb{x},t\right)$ of spatial coordinates vector $\vb{x}$ and time $t$, the heat equation is defined in \cref{eq:heat_equation}.
\begin{equation} \label{eq:heat_equation}
    \pdv{u}{t}=\alpha\laplacian u
\end{equation}
The Laplacian $\laplacian$ denotes sum of all second spatial derivatives. In three dimensions this is $\laplacian\cdot=\pdv{\cdot}{x_1}+\pdv{\cdot}{x_2}+\pdv{\cdot}{x_3}$. The constant $\alpha$ denotes how fast the heat spreads. Intuitively this equation says that the rate at which the temperature changes in time $\pdv{u}{t}$ is proportional to how different the temperature of a spot in the pan with its surroundings. One example of an insight this gives is that given the stove heat stays constant the temperature will no longer change for a particular point once the temperature difference is constant or in other words the function $u\left(\vb{x},t\right)$ approaches a linear function in space. Mathematically this is because of the Laplacian being second spatial derivatives. Intuitively, once the temperature distribution approaches linear, the current spot on the pan is outputting as much heat it is getting.

Many other systems such as waves, quantum dynamics, fluid dynamics, elastics, and many more also define systems using PDEs. PDEs are also used outside of the physical sciences. In Finance, the Black-Scholes-Merton or Black-Scholes equation models the dynamics of the financial market. In ecology, PDEs are used to model population growth which is usefull for modelling location dependent carrying capacity. This is used to model species distribution which can help develop better conservation policies. In the social sciences, PDEs such as the cross-diffusion model is used to explore the evolution of the urban environment <https://doi.org/10.1038/s41598-023-38979-w>. Crowd dynamics is another field where researchers have found it useful to model using PDEs <https://doi.org/10.1016/S0378-4754(00)00228-7, https://doi.org/10.1109/TSMC.2015.2389763>. These are used to model dynamics such as shock-waves, pathing, and crowd flow as a whole. There are a variety of applications, ranging from brid flocks, pedestrian crowds, to robotic swarms <https://doi.org/10.1146/annurev-control-060822-123629>.

Traditional numerical solvers used to solve PDEs have enjoyed many decades of development due to their long history. There are many general approaches to solve PDEs and many more very specific approaches. Some widely used approaches include Finite Element Methods (FEM), Finite Difference Methods (FDM), and Spectral Methods. FEM and FDM both are mesh based, meaning they rely on discretization of the computational domain <ADD DIAGRAM FOR THIS>. There are several challenges associated with this. First, irregular domains such as bone, spider silk, gravel, or other bio-inspired/aggregate materials,  pose a challenge due to the complexity of the domain geometry <https://doi.org/10.1016/0955-7997(91)90014-K, https://doi.org/10.1016/j.jcp.2007.03.025, https://doi.org/10.1038/s41467-024-47831-2>. Another issue is large deformations or mesh entanglement cause these methods to become ineffective <10.1061/(ASCE)EM.1943-7889.0001176>. While there are strategies to mitigate this, by deffinition they are an additional layer of difficulty to the process. Second, multi-scale applications where the micro and macro scales are both important require fine meshes such that the small scale structures are addequately simulated. This creates very large meshes that are very resource intensive <https://doi.org/10.1016/j.jcp.2007.03.025>. Therefore, in problems where these issues important, meshfree methods are prefered. One approach are Spectral methods which utilize basis functions to approaximate a global solution. A popular choice especially problems with known periodicity, are the Fourier Series. Another drawback of traditional methods are the requirement that PDEs and equations to solve be known analytically. This is because the solvers use the equations to formulate the solution. A dealbreaker like this makes traditional numerical approaches unsuited to data dominant problems. Third, a single evaluation of traditional solvers may not be very costly, multiple evaluations can add up. This is very apparent in inverse problems and other applications where multiple evaluations of the solver is required.

With the increasing prevalance of machine learning methods and their use in more and more fields, research into their use for scientific computing has taken off in recent years. While statistical modeling has already been widely used in areas such as physical constants, stellar population studies, risk assesments of events such as earthquakes and coronal mass ejections <geophysics constants: 10.1029/2002JD002865, CME: https://doi.org/10.1016/j.asr.2005.04.089, asteroid: https://doi.org/10.1111/risa.12453, physics constants: https://doi.org/10.1103/RevModPhys.75.403, http://dx.doi.org/10.1214/074921706000000419, stellar population: https://doi.org/10.1093/mnras/stac3287>, the most widespread approach for forward modeling or inverse modeling are physics based numerical models. With the rise of machine learning methods research into their use for solving PDEs have also expanded.

Another aproach to this starts by viewing PDEs as operators, which are generalized mappings between spaces. Familiar operators are functions which are mappings spaces of scalar values or vector values. PDEs on the other hand are operators that map between function spaces. A simple example is the derivative. It takes in a function and returns the derivative of said function. In other words it is an operator that maps between the space of all functions to the space of derivatives of those functions. Another way to view this is by imagining functions as infinite dimensional vectors. Where elements in the vector are the function's value evaluated at every point. The operator can be seen as vector functions mapping between these infinite dimensional vector spaces. Operators are important because a field of research has sprung up around this mathematical concept. Operator learning is the use of machine learning to learn operators from a data driven approach. As an analogy, regression traditionally has been used to approximate the mapping of functions. In the case of operator learning, a mapping between functions may be approximated.
